---
title: "Multi-Tumour Analysis and Sensitivity Analyses with seerTP"
author: "Sameer Mansoori, Shubhram Pandey, Rashi Rani, Barinder Singh, Murat Kurt"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Tumour Analysis and Sensitivity Analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
options(scipen = 999)
```

## Overview

This vignette demonstrates the **batch analysis workflow** for all 11 cancer
types in the bundled `tumour_data_seer` dataset across all supported time
horizons (`"lifetime"`, `"20y"`, `"5y"`) and both information criteria
(`"AIC"`, `"BIC"`).

```{r libs, message = FALSE}
library(seerTP)
library(dplyr)
library(tidyr)
library(ggplot2)
```

---

## Load bundled datasets

```{r data}
data(lifetable_seer)
data(tumour_data_seer)

dplyr::glimpse(tumour_data_seer)
```

---

## Run all scenarios

```{r batch, message = FALSE, warning = FALSE, cache = TRUE}
scenarios <- c("lifetime", "20y", "5y")
criteria  <- c("AIC", "BIC")

run_batch <- function(scen, crit) {
  tumour_data_seer |>
    group_by(Tumor) |>
    summarise(
      out = list(
        run_tumour_analysis(
          tumour       = first(Tumor),
          surv_vec_LR  = c(S0[Recurrence == "LR"], S1[Recurrence == "LR"],
                           S2[Recurrence == "LR"], S3[Recurrence == "LR"],
                           S4[Recurrence == "LR"], S5[Recurrence == "LR"]),
          surv_vec_DR  = c(S0[Recurrence == "DR"], S1[Recurrence == "DR"],
                           S2[Recurrence == "DR"], S3[Recurrence == "DR"],
                           S4[Recurrence == "DR"], S5[Recurrence == "DR"]),
          N_L          = first(N_LR),
          N_D          = first(N_DR),
          prop_male_LR = prop_male[Recurrence == "LR"],
          mean_age_LR  = mean_age[Recurrence == "LR"],
          prop_male_DR = prop_male[Recurrence == "DR"],
          mean_age_DR  = mean_age[Recurrence == "DR"],
          lifetable    = lifetable_seer,
          scenario     = scen,
          criterion    = crit
        )
      ),
      .groups = "drop"
    ) |>
    unnest(out) |>
    mutate(scenario = scen, criterion = crit)
}

all_results <- bind_rows(
  lapply(scenarios, function(s)
    lapply(criteria, function(c) run_batch(s, c)) |> bind_rows()
  )
)

dim(all_results)
```

---

## Base-case results (lifetime horizon, AIC)

```{r basecase}
base <- all_results |>
  filter(scenario == "lifetime", criterion == "AIC") |>
  select(Tumor, P_LL, P_DL_approach2, P_Death_L_approach2, M_months)

knitr::kable(base, digits = 5,
             caption = "Base-case monthly transition probabilities (lifetime, AIC)")
```

---

## Sensitivity: P(DR) by scenario and criterion

```{r sens-plot, fig.cap = "P(DR) across scenarios and criteria"}
all_results |>
  filter(criterion == "AIC") |>
  mutate(scenario = factor(scenario, levels = c("5y", "20y", "lifetime"))) |>
  ggplot(aes(x = P_DL_approach2, y = reorder(Tumor, P_DL_approach2),
             colour = scenario, shape = scenario)) +
  geom_point(size = 3) +
  labs(
    title    = "P(DR) by Cancer Type and Time Horizon",
    subtitle = "AIC-weighted model blend",
    x        = "Monthly P(LR → DR)",
    y        = NULL,
    colour   = "Horizon",
    shape    = "Horizon"
  ) +
  scale_colour_brewer(palette = "Set1") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

---

## AIC vs BIC comparison (lifetime horizon)

```{r aic-bic, fig.cap = "AIC vs BIC — P(DR)"}
all_results |>
  filter(scenario == "lifetime") |>
  select(Tumor, criterion, P_DL_approach2) |>
  pivot_wider(names_from = criterion, values_from = P_DL_approach2) |>
  ggplot(aes(x = AIC, y = BIC, label = Tumor)) +
  geom_point(size = 3, colour = "#E41A1C") +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 20) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  labs(
    title = "AIC- vs BIC-weighted P(DR) — Lifetime Horizon",
    x     = "P(LR → DR) under AIC weighting",
    y     = "P(LR → DR) under BIC weighting"
  ) +
  theme_minimal(base_size = 11)
```

> **Note:** `ggrepel` is used for label placement above.  Install with
> `install.packages("ggrepel")` if needed.

---

## Exporting results

```{r export, eval = FALSE}
# Export combined results to CSV
write.csv(
  all_results,
  file      = "seerTP_all_results.csv",
  row.names = FALSE
)

# Export base-case to Excel
openxlsx::write.xlsx(
  base,
  file      = "seerTP_basecase.xlsx",
  sheetName = "Lifetime_AIC"
)
```

---

## Session information

```{r session}
sessionInfo()
```
