---
title: "Introduction to seerSurv: Cancer Recurrence Transition Probabilities"
author: "Sameer Mansoori, Shubhram Pandey, Rashi Rani, Barinder Singh, Murat Kurt"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to seerSurv}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%"
)
options(scipen = 999)
```

## Background

State-transition health economic models of early-stage cancers often include a
**local/regional recurrence (LR)** state that sits between the
disease-free state and distant recurrence (DR) or death.  Populating such a
model requires three monthly transition probabilities from LR:

| Transition | Symbol |
|---|---|
| Remain in LR | $P_{\text{LL}}$ |
| Progress to DR | $P_{\text{DL}}$ |
| Die while in LR | $P_{\text{death} \mid \text{LR}} = 1 - P_{\text{LL}} - P_{\text{DL}}$ |

Trial-derived estimates are rarely available because clinical studies are
typically powered for overall or disease-free survival, and LR follow-up is
too short to observe the full recurrence cascade.

**seerSurv** fills this gap by deriving the three probabilities from aggregate
(population-level) survival data extracted from the
[SEER-17 registry](https://seer.cancer.gov/), covering ten cancer types.

---

## Methodological overview

```{r method-diagram, echo = FALSE, fig.cap = "seerSurv workflow"}
# Simple text diagram rendered as a figure caption placeholder
```

1. **Pseudo-IPD construction** — Aggregate survival proportions at years 0–5
   are converted to pseudo-individual patient data (`prep_ipd()`).

2. **Parametric modelling** — Six distributions (exponential, Weibull,
   Gompertz, log-logistic, log-normal, gamma) are fitted to the pseudo-IPD
   using weighted maximum likelihood via `flexsurv` (`fit_models()`).

3. **Model averaging** — The top-$k$ distributions by AIC or BIC receive
   relative-likelihood weights; the final curve is a convex blend
   (`compute_weights()`, `blend_survival()`).

4. **Background mortality adjustment** — Disease-specific survival is
   multiplied by age- and sex-specific background survival from the US-CDC
   life-table to obtain net (all-cause) survival (`make_background_surv()`).

5. **Sojourn time and $P_{\text{LL}}$** — The mean LR sojourn time $M$
   (months) equals the area difference between the net LR and net DR curves.
   $P_{\text{LL}}$ solves $\sum_{t=0}^{N} P_{\text{LL}}^t = M$.

6. **Convolution for $P_{\text{DL}}$** — The modelled LR survival is
   approximated as the sum of patients still in LR and those who have
   transitioned to DR.  $P_{\text{DL}}$ is estimated by least-squares
   (`run_tumour_analysis()`).

---

## Quickstart: single tumour type

```{r melanoma-demo, message = FALSE}
library(seerSurv)
library(dplyr)

data(lifetable_seer)

# Five-year SEER-17 survival proportions — Melanoma
s_lr <- c(1, 0.9959, 0.9886, 0.9831, 0.9783, 0.9754)
s_dr <- c(1, 0.5491, 0.4396, 0.3945, 0.3630, 0.3410)

result <- run_tumour_analysis(
  tumour       = "Melanoma",
  surv_vec_LR  = s_lr,
  surv_vec_DR  = s_dr,
  N_L          = 64775L,
  N_D          = 3121L,
  prop_male_LR = 0.580,
  mean_age_LR  = 61,
  prop_male_DR = 0.692,
  mean_age_DR  = 62,
  lifetable    = lifetable_seer,
  scenario     = "lifetime",
  criterion    = "AIC"
)

knitr::kable(result, digits = 5,
             caption = "Monthly transition probabilities — Melanoma (lifetime, AIC)")
```

---

## Step-by-step walkthrough

### Step 1: Prepare pseudo-IPD

```{r ipd}
t    <- 0:5
ipd_L <- prep_ipd(t, s_lr)
ipd_D <- prep_ipd(t, s_dr)
ipd_L
```

### Step 2: Fit parametric models

```{r fit, message = FALSE}
mods_L <- fit_models(ipd_L[-1, ])   # drop time = 0 row
mods_D <- fit_models(ipd_D[-1, ])

names(mods_L)
```

### Step 3: Compute AIC-based weights

```{r weights}
ic_L   <- extract_ic(mods_L, "AIC")
wts_L  <- compute_weights(ic_L, top_k = 3, criterion = "AIC")
wts_L
```

### Step 4: Blend survival curves

```{r blend}
grid     <- seq(0, 39, by = 1)   # 39-year lifetime horizon from age 61
S_blend_L <- blend_survival(mods_L, wts_L, grid)
head(S_blend_L)
```

### Step 5: Adjust for background mortality

```{r bgmort}
lt <- lifetable_seer |>
  mutate(
    btrate_LR = Males * 0.580 + Females * 0.420,
    btrate_DR = Males * 0.692 + Females * 0.308
  )

B_LR <- make_background_surv(lt, mean_age = 61, rate_col = "btrate_LR",
                              time_grid_years = grid)
B_DR <- make_background_surv(lt, mean_age = 62, rate_col = "btrate_DR",
                              time_grid_years = grid)
head(B_LR)
```

### Step 6: Visualise

```{r plot, fig.cap = "Blended and net survival curves — Melanoma"}
ic_D  <- extract_ic(mods_D, "AIC")
wts_D <- compute_weights(ic_D, top_k = 3, criterion = "AIC")
S_blend_D <- blend_survival(mods_D, wts_D, grid)

U_L <- left_join(S_blend_L, B_LR, by = "time") |>
  mutate(surv = surv.x * surv.y) |> select(time, surv)

U_D <- left_join(S_blend_D, B_DR, by = "time") |>
  mutate(surv = surv.x * surv.y) |> select(time, surv)

plot_survival_curves(S_blend_L, S_blend_D, U_L, U_D, tumour = "Melanoma")
```

---

## Multi-tumour analysis using bundled data

```{r multi, message = FALSE, warning = FALSE}
data(tumour_data_seer)

results_multi <- tumour_data_seer |>
  group_by(Tumor) |>
  summarise(
    out = list(
      run_tumour_analysis(
        tumour       = first(Tumor),
        surv_vec_LR  = c(S0[Recurrence == "LR"], S1[Recurrence == "LR"],
                         S2[Recurrence == "LR"], S3[Recurrence == "LR"],
                         S4[Recurrence == "LR"], S5[Recurrence == "LR"]),
        surv_vec_DR  = c(S0[Recurrence == "DR"], S1[Recurrence == "DR"],
                         S2[Recurrence == "DR"], S3[Recurrence == "DR"],
                         S4[Recurrence == "DR"], S5[Recurrence == "DR"]),
        N_L          = first(N_LR),
        N_D          = first(N_DR),
        prop_male_LR = prop_male[Recurrence == "LR"],
        mean_age_LR  = mean_age[Recurrence == "LR"],
        prop_male_DR = prop_male[Recurrence == "DR"],
        mean_age_DR  = mean_age[Recurrence == "DR"],
        lifetable    = lifetable_seer,
        scenario     = "lifetime",
        criterion    = "AIC"
      )
    ),
    .groups = "drop"
  ) |>
  tidyr::unnest(out)

knitr::kable(
  results_multi |> select(Tumor, P_LL, P_DL_approach2, P_Death_L_approach2,
                           M_months),
  digits  = 5,
  caption = "Monthly transition probabilities across 11 cancer types (lifetime, AIC)"
)
```

---

## Sensitivity analysis: time horizons and model-selection criteria

```{r sensitivity, message = FALSE, warning = FALSE}
scenarios  <- c("lifetime", "20y", "5y")
criteria   <- c("AIC", "BIC")

sens_grid <- expand.grid(scenario = scenarios, criterion = criteria,
                         stringsAsFactors = FALSE)

sens_results <- lapply(seq_len(nrow(sens_grid)), function(i) {
  run_tumour_analysis(
    tumour       = "Melanoma",
    surv_vec_LR  = s_lr,
    surv_vec_DR  = s_dr,
    N_L          = 64775L,
    N_D          = 3121L,
    prop_male_LR = 0.580,
    mean_age_LR  = 61,
    prop_male_DR = 0.692,
    mean_age_DR  = 62,
    lifetable    = lifetable_seer,
    scenario     = sens_grid$scenario[i],
    criterion    = sens_grid$criterion[i]
  ) |>
    mutate(scenario  = sens_grid$scenario[i],
           criterion = sens_grid$criterion[i])
})

sens_df <- bind_rows(sens_results) |>
  select(scenario, criterion, P_LL, P_DL_approach2, P_Death_L_approach2)

knitr::kable(sens_df, digits = 5,
             caption = "Sensitivity analysis — Melanoma across scenarios and criteria")
```

---

## Interpreting the outputs

| Column | Interpretation |
|---|---|
| `P_LL` | Per-month probability of remaining in LR state |
| `P_DL_approach2` | Per-month probability of progressing from LR to DR |
| `P_Death_L_approach2` | Per-month probability of dying while in LR |
| `RMST_L_years` | Restricted mean survival time — LR population (net, years) |
| `RMST_D_years` | Restricted mean survival time — DR population (net, years) |
| `M_years` / `M_months` | Mean sojourn time in LR state |

These probabilities can be plugged directly into a Markov cohort model or a
partitioned survival model that includes an LR health state.

---

## References

Mansoori S, Pandey S, Rani R, Singh B, Kurt M (2025).
"Deriving Cancer Progression Rates After Local/Regional Recurrence Using
Aggregate Survival Data: An R Package for Health Economic Evaluations."
*Value in Health* (submitted).

SEER Program. National Cancer Institute. <https://seer.cancer.gov/>.

Jackson C (2016). "flexsurv: A Platform for Parametric Survival Modelling
in R." *Journal of Statistical Software*, **70**(8), 1--33.
doi:[10.18637/jss.v070.i08](https://doi.org/10.18637/jss.v070.i08).
